SOURCE SEPARATION METHODS FOR COMPUTER-ASSISTED ORCHESTRATION

### 2021-06-15 source_separation/paper/paper.tex
Add qualitative evaluation section

	Orchestration~\cite{adler1989study} is hard to define, however one could say that orchestration is the set of musical writing techniques that focus on timbre, or \textit{tone color} of the sound. A simplistic  definition of musical orchestration would be to say that it consists of the distribution of the different musical voices of a piece to the instruments of the orchestra. Orchestration has been traditionally taught in an intuitive and non-formalized way, and computer-assisted orchestration seeks to help composers by accelerating certain creative processes.

@@

%  The codebase for this paper can be found at: \url{https://github.com/dzluke/MLSP2021}, and you can listen to a selection of targets and orchestrations here: \url{https://dzluke.github.io/MLSP2021/}.
  The codebase for this paper can be found at: \url{anonymous}, and a selection of targets and orchestrations are available for listenig at: \url{anonymous}.

@@

  \begin{figure*}[t]
%  \begin{figure}[t]
    \centering
      \includegraphics[width=0.7\textwidth]{figures/diagram.png}
      \caption{Diagram comparing the process of orchestration with and without separation. At the top, the full target orchestration is created simply by orchestrating the target. At the bottom, the separation process is performed, and the resulting sub-targets are individually orchestrated with subsets of the orchestra before being recombined to obtain the full orchestration.}\label{fig:full_diagram}
%  \end{figure}  
  \end{figure*}

@@

    \subsection{Testing}\label{subsec:testing}
    In order to compare the effectiveness of the separation methods, we compare them with \textit{a full target orchestration}. The \textbf{full target orchestration} is the orchestration of the target without any separation performed.

    The procedure we use for testing is as follows:
    \begin{enumerate}
      \item The target is created as a combination of four randomly chosen sources, which are offset to begin playing at different times.
      \item The target, without any separation performed, is orchestrated using the entire orchestra. This is called the full target orchestration.
      \item Each separation method is performed on the target, each splitting the target into four sub-targets. 
      \item The full orchestra is randomly split into four equal sized subsets, each containing 7 to 8 instruments.
      \item For each method, the four sub-targets are separately orchestrated, each using a different subset of the orchestra, and then each orchestration is combined to play simultaneously. This creates the \textbf{separated orchestrations}.
      \item The distances between target and orchestration are calculated for the full target orchestration and the separated orchestration.
    \end{enumerate}

@@

  \section{Discussion}\label{sec:discussion}
  In this paper, we addressed the problem of using source separation as a pre-processing step for computer-assisted orchestration. The results show that performing source separation followed by a static orchestration of each individual sub-target improved the spectral distance between the original target and its orchestrated version in three out of five of the methods tested. We believe this shows that source separation is a useful pre-processing stage for computer-assisted orchestration. We also conducted a qualitative assessment of the orchestrations by acoustic inspection and found that the separated orchestrations are better from a musical standpoint.

  Furthermore, the one unsupervised method we tested, NMF, outperformed all the supervised methods. This suggests that methods like NMF can still be useful for specific problems when available data is limited. It is logical that unsupervised methods will perform well for problems in which there is limited data, as they require less data than supervised methods. It is important to notice that for the supervised methods, we used pre-trained versions of the neural models. All of these models, except for TDCN++, were trained on musical data. However, we tested them on a wide range of targets, many of which were not strictly musical. This could explain why Demucs, Open-Unmix, and TDCN performed worse than NMF, and why TDCN++, which was trained on arbitrary sounds, outperformed the other supervised methods. Therefore, we may infer from our results that models trained on generic sounds generate a better separation for orchestration when compared to models trained only on musical sounds.

  Since unsupervised methods perform well, it is reasonable to think that supervised methods trained on an appropriate dataset of arbitrary sounds could yield better overall performance for orchestration.

>>>>

	Orchestration~\cite{Adler2016} is hard to define, however one could say that orchestration is the set of musical writing techniques that focus on timbre, or \textit{tone color} of the sound. A simplistic  definition of musical orchestration would be to say that it consists of the distribution of the different musical voices of a piece to the instruments of the orchestra. Orchestration has been traditionally taught in an intuitive and non-formalized way, and computer-assisted orchestration seeks to help composers by accelerating certain creative processes.

@@

%  The codebase for this paper can be found at: \url{https://github.com/dzluke/MLSP2021}, and you can listen to a selection of targets and orchestrations here: \url{https://dzluke.github.io/MLSP2021/}.
  The codebase for this paper can be found at: \url{anonymous}, and a selection of targets and orchestrations are available for listening at: \url{anonymous}.

@@

  \begin{figure}[t]
    \centering
      \includegraphics[width=\columnwidth]{figures/diagram.png}
      \caption{Diagram comparing the process of orchestration with and without separation. At the top, the full target orchestration is created simply by orchestrating the target. At the bottom, the separation process is performed, and the resulting sub-targets are individually orchestrated with subsets of the orchestra before being recombined to obtain the full orchestration.}\label{fig:full_diagram}
  \end{figure}

@@

    \subsection{Testing}\label{subsec:testing}
    The procedure we use for testing is as follows:
    \begin{enumerate}
      \item The target is created as a combination of four randomly chosen sources, which are offset to begin playing at different times.
      \item The target, \textit{without any separation performed}, is orchestrated using the entire orchestra. This creates what we call the \textit{full target orchestration}.
      \item Each separation method is performed on the target, each splitting the target into four sub-targets. 
      \item The full orchestra is randomly split into four equal sized subsets, each containing 7 to 8 instruments.
      \item For each method, the four sub-targets are separately orchestrated, each using a different subset of the orchestra, and then each orchestration is combined to play simultaneously. This creates the \textit{separated orchestrations}.
      \item The distances between target and orchestration are calculated for the full target orchestration and the separated orchestration.
    \end{enumerate}

@@

  \section{Discussion}\label{sec:discussion}
  In this paper, we addressed the problem of using source separation as a pre-processing step for computer-assisted orchestration. The results show that performing source separation followed by a static orchestration of each individual sub-target improved the spectral distance between the original target and its orchestrated version in three out of five of the methods tested. We believe this shows that source separation is a useful pre-processing stage for computer-assisted orchestration.

  In order to fully understand the impact of source separation on the orchestration process, we also conducted a qualitative evaluation of the generated orchestrations by acoustic inspection.
This evaluation showed that source separation significantly improves the quality or orchestrations from a musical standpoint. The next paragraphs will try to explain this in more detail.

	A major disadvantage of the orchestrations generated without any source separation is that, when two or more sources overlap in time, the orchestration algorithm cannot distinguish between them. As such, the orchestration appears to be more rigid: the whole orchestra suddenly changes configuration when a new source dominates, without any blending with the previous dominating source. This is against a well know orchestration principle called \emph{dovetailing} \cite[pp. 467-473]{Adler2016}. Dovetailing is a well known practice aimed at creating a blend of the instruments used during orchestration, in order to transition between timbres. Technically, dovetailing is the swapping and overlapping of musical lines between different instruments: musical lines are scattered across multiple instruments, connected with overlapping pitches.

	An interesting result that we discovered after acoustic inspection is that the orchestrations generated after source separation tend to favor dovetailing. Since only a portion of the orchestra is assigned to each source and since the sources are inherently asynchronous with each other, the final orchestration is more fluid and appears more interesting, from a musical standpoint.

  Quantitatively, the one unsupervised method we tested, NMF, outperformed all the supervised methods. This suggests that methods like NMF can still be useful for specific problems when available data is limited. It is logical that unsupervised methods will perform well for problems in which there is limited data, as they require less data than supervised methods. It is important to notice that for the supervised methods, we used pre-trained versions of the neural models. All of these models, except for TDCN++, were trained on musical data. However, we tested them on a wide range of targets, many of which were not strictly musical. This could explain why Demucs, Open-Unmix, and TDCN performed worse than NMF, and why TDCN++, which was trained on arbitrary sounds, outperformed the other supervised methods. Therefore, we may infer from our results that models trained on generic sounds generate a better separation for orchestration when compared to models trained only on musical sounds. Since unsupervised methods perform well, it is reasonable to think that supervised methods trained on an appropriate dataset of arbitrary sounds could yield better overall performance for orchestration.
